---
title: "Data Cleaning Practice"
author: "PM"
date: "2023-07-31"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning Practice Markdown

This is a R markdown of important data cleaning code for future projects. For more information on creating a markdown visit: <http://rmarkdown.rstudio.com>.

## Important packages

```{r common packags for data cleaning}
library("tibble")
library("tidyr")
library("ggplot2")
library("eeptools")
print("Done")
```

## Getting the data

```{r get the data}
library(readr)
my_fake_demographics <- read_csv("my_fake_demographics.csv")
View(my_fake_demographics)
dem.df2 <- my_fake_demographics # Keep the name simple 
head(dem.df2)
```

```{r get to know the data1}
summary(dem.df2)
```

```{r get to know the data2}
str(dem.df2)
```

```{r get to know the data3}
names(dem.df2)
```

I can see the dimensions, categories, values, and stats of the data

## Missing values

```{r finding missing values}
sapply(dem.df2, function(x) sum(is.na(x)))
```

```{r missing value map}
library("visdat")
vis_miss(dem.df2) 
```

## Imputation

Now lets correct the missing values in "RACE", "ETHNICITY", and "SEX".

```{r imputation}
dem.df2[!complete.cases(dem.df2),] # show rows with missing values
```

# Ttanic data

These are the packages needed for the next part (imputation)

```{r mice}
library(ggplot2)
library(dplyr)
library(titanic)
library(cowplot)

titanic_train$Age

```

```{r mice getting the data}
sapply(titanic_train, function(x) sum(is.na(x)))
library("visdat")
vis_miss(titanic_train)
```

# Simple Imputation

The value_imputed variable will store a data.frame of the imputed ages. The imputation itself boils down to replacing a column subset that has a value of NA with the value of our choice. This will be:

Zero: constant imputation, feel free to change the value. Mean (average): average age after when all NA's are removed. Median: median age after when all NA's are removed. \## Visualize variable distribution This step is important to compare variables before and after imputation.

```{r}
ggplot(titanic_train, aes(Age)) +
  geom_histogram(color = "#000000", fill = "#0099F8") +
  ggtitle("Variable distribution") +
  theme_classic() +
  theme(plot.title = element_text(size = 18))
```

```{r simple imputation1}
value_imputed <- data.frame(
  original = titanic_train$Age,
  imputed_zero = replace(titanic_train$Age, is.na(titanic_train$Age), 0),
  imputed_mean = replace(titanic_train$Age, is.na(titanic_train$Age), mean(titanic_train$Age, na.rm = TRUE)),
  imputed_median = replace(titanic_train$Age, is.na(titanic_train$Age), median(titanic_train$Age, na.rm = TRUE))
)
value_imputed
```

### New variable distribution

```{r simple imputation2}
h1 <- ggplot(value_imputed, aes(x = original)) +
  geom_histogram(fill = "#ad1538", color = "#000000", position = "identity") +
  ggtitle("Original distribution") +
  theme_classic()
h2 <- ggplot(value_imputed, aes(x = imputed_zero)) +
  geom_histogram(fill = "#15ad4f", color = "#000000", position = "identity") +
  ggtitle("Zero-imputed distribution") +
  theme_classic()
h3 <- ggplot(value_imputed, aes(x = imputed_mean)) +
  geom_histogram(fill = "#1543ad", color = "#000000", position = "identity") +
  ggtitle("Mean-imputed distribution") +
  theme_classic()
h4 <- ggplot(value_imputed, aes(x = imputed_median)) +
  geom_histogram(fill = "#ad8415", color = "#000000", position = "identity") +
  ggtitle("Median-imputed distribution") +
  theme_classic()

plot_grid(h1, h2, h3, h4, nrow = 2, ncol = 2)
```

## MICE

MICE stands for Multivariate Imputation via Chained Equations, and it's one of the most common packages for R users. It assumes the missing values are missing at random (MAR). The basic idea behind the algorithm is to treat each variable that has missing values as a dependent variable in regression and treat the others as independent (predictors). The R mice packages provide many univariate imputation methods.

[https://appsilon.com/imputation-in-r/#:\~:text=Impute%20Missing%20Values%20in%20R%20with%20MICE&text=It%20assumes%20the%20missing%20values,others%20as%20independent%20(predictors).](https://appsilon.com/imputation-in-r/#:~:text=Impute%20Missing%20Values%20in%20R%20with%20MICE&text=It%20assumes%20the%20missing%20values,others%20as%20independent%20(predictors).)

We will be using the titanic data to learn about imputation package. \### Examine the data It is important to get a visual representation of missing data and variable distribution when examining new data.

```{r mice1}
library(mice)

titanic_numeric <- titanic_train %>%
  select(Survived, Pclass, SibSp, Parch, Age)

md.pattern(titanic_numeric)
```

I prefer the \`sapply\` method. I shows the missing values clearly.

```{r sapply}
sapply(titanic_numeric, function(x) sum(is.na(x)))
```

Onto the imputation now. We'll use the following MICE imputation methods:

-   pmm: Predictive mean matching.
-   cart: Classification and regression trees.
-   laso.norm: Lasso linear regression.

```{r mice2}
mice_imputed <- data.frame(
  original = titanic_train$Age,
  imputed_pmm = complete(mice(titanic_numeric, method = "pmm"))$Age,
  imputed_cart = complete(mice(titanic_numeric, method = "cart"))$Age,
  imputed_lasso = complete(mice(titanic_numeric, method = "lasso.norm"))$Age
)
mice_imputed 
```

### Examine the new data

```{r mice3}
h1 <- ggplot(mice_imputed, aes(x = original)) +
  geom_histogram(fill = "#ad1538", color = "#000000", position = "identity") +
  ggtitle("Original distribution") +
  theme_classic()
h2 <- ggplot(mice_imputed, aes(x = imputed_pmm)) +
  geom_histogram(fill = "#15ad4f", color = "#000000", position = "identity") +
  ggtitle("PMM-imputed distribution") +
  theme_classic()
h3 <- ggplot(mice_imputed, aes(x =  imputed_cart)) +
  geom_histogram(fill = "#1543ad", color = "#000000", position = "identity") +
  ggtitle("CART-imputed distribution") +
  theme_classic()
h4 <- ggplot(mice_imputed, aes(x = imputed_lasso)) +
  geom_histogram(fill = "#ad8415", color = "#000000", position = "identity") +
  ggtitle("Lasso-imputed distribution") +
  theme_classic()

plot_grid(h1, h2, h3, h4, nrow = 2, ncol = 2)
```

Note that there are negative ages that show in the lasso results. negative values will need to be corrected manually.

There are many ways to impute in R. Check out [[https://appsilon.com/imputation-in-r/#:\\\\](https://appsilon.com/imputation-in-r/#:\\){.uri}]([https://appsilon.com/imputation-in-r/#:\\](https://appsilon.com/imputation-in-r/#:\){.uri}){.uri}\~:text=Impute%20Missing%20Values%20in%20R%20with%20MICE&text=It%20assumes%20the%20missing%20values,others%20as%20independent%20(predictors) for more information. \# Patient demographics data practice Now back to working with our patient demographic data `dem.df2` (these are fake demographics).

```{r dem.df2}
head(dem.df2)
```

## Add CURRENT_AGE column

The objective of this task is to learn how to copy, add, and calculate a new data to a df.

```{r add col}
dem.df2$CURRENT_AGE <- 0 # add the CURRENT_AGE column with 0 as a place holder value
head(dem.df2)
```

```{r age}
# create the date variable
as.Date
```
``` {r drop na}
install.packages("tidyr")
library("tidyr")
dem.df2 %>% drop_na()
```
```{r age2}
# solution: Missing value where true/false needed
library("eeptools")
dem.df2$CURRENT_AGE <- dem.df2$PAT_DOB # copy a column
library(lubridate)
# make sure that the dates are in the same format
dem.df2$CURRENT_AGE <- as.Date(dem.df2$PAT_DOB, format = "%m/%d/%Y") 
dem.df2$PAT_DOB <- as.Date(dem.df2$PAT_DOB, format = "%m/%d/%Y")
# convert to age in years 
dem.df2$CURRENT_AGE <- age_calc(dem.df2$CURRENT_AGE, units = "years")
head(dem.df2)
```

# Pulling specific data from a df

Extract all female patients ages greater than or equal to 10.

```{r data pull}
subset(dem.df2, SEX == 'Female' & CURRENT_AGE >= 10)
```

# Plots

Now we will work on creating plots. Plot patient "SEX".Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. There are more complicated plot in the 'Simple Imputation' section.

```{r plots}
library("tidyverse")
```

```{r plot}
dem.df2 %>% count(SEX) 
ggplot(data = dem.df2) + 
  geom_bar(mapping = aes(x = SEX)) # simple bar graph
```

# Exporting data

This is the template to export data. `write.csv(DataFrame Name, "Path to export the DataFrame\\File Name.csv", row.names=FALSE)`

```{r export}
# Load readr package
library("readr")
```

```{r export2}
write_csv(dem.df2, "new_patientdem.csv") 
```

# Resources

<https://www.r-bloggers.com/2021/04/handling-missing-values-in-r/> <https://stackoverflow.com/questions/27096485/change-a-column-from-birth-date-to-age-in-r> 
<[https://sparkbyexamples.com/r-programming/r-subset-data-frame-with-examples/#:\~:text=If%20you%20wanted%20to%20get,variables)%20from%20the%20data%20frame.]>

<(https://sparkbyexamples.com/r-programming/r-subset-data-frame-with-examples/#:~:text=If%20you%20wanted%20to%20get,variables)%20from%20the%20data%20frame.){.uri}>

<https://michaelgastner.com/R_for_QR/extracting-values-from-data-frames.html> <https://community.rstudio.com/t/dplyr-way-s-and-base-r-way-s-of-creating-age-group-from-age/89226/3> <https://www.datanovia.com/en/lessons/subset-data-frame-rows-in-r/> <https://r4ds.had.co.nz/data-visualisation.html> <https://www.infoworld.com/article/3573577/how-to-count-by-groups-in-r.html> <https://dplyr.tidyverse.org/reference/mutate.html>

<https://sparkbyexamples.com/r-programming/r-export-csv-using-write-csv/>
